\Title

### 2022-09-16 Grimaldi-etal-BiolCybernetics.tex
figures

In this work, we study the emergence of such spatio-temporal spiking motifs when training a single layer of spiking neurons on a supervised classification task (\seeFig{motion-task}). We develop a SNN-like method able to learn hetero-synaptic delays to perform motion detection on a synthetic event-based dataset. Because neuromorphic devices are, by design, good candidates to integrate computations with time, we highlight the fact that this event-driven algorithm is transferable to such hardware.
%
\section{Methods}
\label{sec:methods}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
Let us now formally define the hetero-synaptic model, as well as the task and how we will solve it. First, we will define the model for supervised learning of weights and delays in a feed-forward layer of a spiking neural network (SNN). Our SNN model will be tested on a computationally relevant task of motion detection in an event-based sensory input. By analogy with biological conditions, we will design a paradigm simulating eye movements where an input image is translated by a parameterized saccadic trajectory. The task is then to guess the motion as accurately as possible (see \seeFig{motion-task}). This is, for example, useful in real-life situations to compensate for eye, head or body movements and to provide robust image categorization by stabilizing the image on the retina. To this end, we will extend the SNN model to be efficient for the resolution of the task by including convolutional kernels. We will finally describe its implementation on conventional computers. 
%
\subsection{Hetero-synaptic delays model as a Logistic Regression with a temporal convolution}
%%%-----------------------------------------------------------------
Let's first define a layer of hetero-synaptic spiking neurons $\postsynaddr \in \postsynaddrspace$ by first describing how each neuron connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses, and each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. For each neuron $\postsynaddr$, we thus define a set of $\Nsyn^\postsynaddr$ synapses, as  $\synapse^\postsynaddr = \{(\presynaddr^\postsynaddr_\ranksyn, \synapticweight^\postsynaddr_\ranksyn, \synapticdelay^\postsynaddr_\ranksyn)\}_{\ranksyn \in [1,\Nsyn^\postsynaddr]}$, where each synapse $\synapse^\postsynaddr_\ranksyn$ is associated to a weight $\synapticweight^\postsynaddr_\ranksyn$, a delay $\synapticdelay^\postsynaddr_\ranksyn$ and a presynaptic address $\presynaddr^\postsynaddr_\ranksyn$. This defines the full connectivity of the hetero-synaptic model.

@@

The activation function of our spiking neuron is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
It transforms this list of weights into a probability with the following formula:
$
Pr(k=\postsynaddr \; \vert \; \timev) =
\frac 1 Z
{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
$ 
where $\mathcal{C}^\postsynaddr(\timev) = \sum
\activeweights^\postsynaddr(t)
$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%
In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the hetero-synaptic delays model. 
%
\subsection{Task definition: fast motion detection}
%%%-----------------------------------------------------------------
Let us now define a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the field of vision. In animals with a fovea, this is particularly useful as it allows to move the area with the highest density of photoreceptors in the environment, for example at a point of interest. For example, saccades are rapid movements of the eyeball that reposition the center of vision. In humans, these are very frequent (on average one every 3 per second~\citep{dandekar_neural_2012}). They are produced very rapidly (about $80~\ms$) and display a whole range of speeds. At a more microscopic scale, the human gaze moves with an incessant drift similar to a Brownian-like trajectory~\citep{poletti_head-eye_2015}. To maintain the full generality of the task, we will define eye movements using a form of random walk~\citep{engbert_integrated_2011}. This approach first defines a finite set of possible 2D motions in polar coordinates. Based on the distribution of biological motions, we simplified it by selecting a set of eye movements as the Cartesian product of N-V-phi=8 linearly spaced motion directions and N-Y-velocity=7 different velocities (see inset of \seeFig{motion-task}-(a)). Note that the velocities are sampled on a geometric scale between $1/4$ and $v-speed_base = 4$. Next, we define one trajectory of gaze as segments whose duration is drawn from a Poisson distribution with an average block length = $20~\ms$ - similar to a Lévy flight~\citep{}. Finally, the trajectory is integrated by assuming 1/ that velocities are uniformly and independently sampled from the set of different motion-sets and 2/ that motion is uniform during a time segment. The resulting instances yield trajectories qualitatively similar to those observed for human eye movements (\seeFig{motion-task}-(a)).

%%%-----------------------------------------------------------------
\begin{figure}[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \note{TODO: redo figure according to caption}
    \caption{
    {\bf Motion detection task.} {\bf (a)} To mimic the effect of a saccadic-like movement, the axis of view moves following a step-like random walk, and we show here on example path of the trajectory. {\bf (b)} We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view (here the cross at the center of the image) such as to reproduce the effect of displacing the eye. {\bf (c)} Recording the dynamics of the sub-image as a function of time, it generates a naturalistic movie which may be transformed to an event-based representation. Mimicking the retina, this representation codes for proportional increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. This will constitute the input to the spiking neural network.}
    \label{fig:motion-task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (see \seeFig{motion-task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration N-T = $250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a movement of one pixel per frame (i.e., per time bin).

@@

To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. If, on a specific pixel at that specific timestamp, the absolute value of this gradient exceeds a threshold, an event is generated. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion-task}-(c)). The goal here is to infer the correct motion solely by observing these events. 

@@

@@

    \vspace{-3cm}

@@

%
%%%-----------------------------------------------------------------
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/accuracy.pdf}
    \caption{Accuracy as a function of the number of computation load for the hetero-synaptic delays model (in blue) and for a method using 2D time surfaces (in red). The relative computational load (on a log axis) is controlled by changing the percentage of active weights relative to the dense convolution kernel. We observe a similar accuracy than HOTS, yet that our model could achieve a similar accuracy with significantly fewer coefficients.}
    \label{fig:accuracy}
\end{figure}
%%%-----------------------------------------------------------------
%
\subsection{Testing with natural-like textures}
%%%-----------------------------------------------------------------
To test our model, we will quantify its ability to categorize different motions. In that order, we use a set of synthetic visual stimuli, \textit{Motion Clouds}~\citep{leon_motion_2012} which are natural-like random textures for which we can control for velocity, among other parameters (\seeFig{motionclouds}). In particular, we will set the spatial size to $(\Nx, \Ny)$ and consider a discretization of time with a time step of $1$ such that $\timev \in \mathbb{N}$. Movies' duration are here set to $\Ntime=400$.
This procedure defines a set of textures with different spatial properties and different motions $\va{v}_k$ with  $1 \le k \le \Nclass$ and $\Nclass=8$ defined by a constant speed and linearly spaced directions:
% of a motion cloud is determined by a constant $k_v$: with a constant speed $\speed$ (see \fig{motionclouds}-(a)): 
%\begin{equation*}
%%\begin{gathered}
%\va{v}_k = %\binom{\cos(\frac{2\pi\cdot k_v}{N_v})}{\sin(\frac{2\pi\cdot k_v}{N_v})}
%  \begin{pmatrix}
%    \cos(2\pi\cdot \frac{k}{\Nclass})\\
%    \sin(2\pi\cdot \frac{k}{\Nclass})
%  \end{pmatrix}
%%\end{gathered}
%\end{equation*}
$
\speed_\kernelind = 
  ( 
    \speed \cdot \cos(2\pi\cdot \frac{\kernelind}{\Nspeed}),
    \speed \cdot \sin(2\pi\cdot \frac{\kernelind}{\Nspeed})
  )
$ (\seeFig{motionclouds}-(a) for an illustration).
For any given velocity, we also varied the parameters of the textures, such as the mean and variance of the orientation or spatial frequency content. This method provides a rich dataset of textured movies for which we know the ground truth for motion.
%
\begin{figure}[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion-clouds.pdf}
    \caption{{\bf Motion detection task.} {\bf (a)} The motion direction represented as the plain red vector, other possible motion directions are represented in light red. {\bf (b)} A screenshot of one generated naturalistic textured stimulus at a specific time. % $\timev_\arank$. 
    {\bf (c)} The corresponding ON (in red) and OFF (in blue) event stream generated from the stimuli on (b) and constituting the input to the spiking neural network.}
    \label{fig:motionclouds}
\end{figure}

>>>>

In this work, we study the emergence of such spatio-temporal spiking motifs when training a single layer of spiking neurons on a supervised classification task (\seeFig{motion_task}). We develop a SNN-like method able to learn hetero-synaptic delays to perform motion detection on a synthetic event-based dataset. Because neuromorphic devices are, by design, good candidates to integrate computations with time, we highlight the fact that this event-driven algorithm is transferable to such hardware.
%
\section{Methods}
\label{sec:methods}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
Let us now formally define the hetero-synaptic model, as well as the task and how we will solve it. First, we will define the model for supervised learning of weights and delays in a feed-forward layer of a spiking neural network (SNN). Our SNN model will be tested on a computationally relevant task of motion detection in an event-based sensory input. By analogy with biological conditions, we will design a paradigm simulating eye movements where an input image is translated by a parameterized saccadic trajectory. The task is then to guess the motion as accurately as possible (see \seeFig{motion_task}). This is, for example, useful in real-life situations to compensate for eye, head or body movements and to provide robust image categorization by stabilizing the image on the retina. To this end, we will extend the SNN model to be efficient for the resolution of the task by including convolutional kernels. We will finally describe its implementation on conventional computers. 
%
\subsection{Hetero-synaptic delays model as a Logistic Regression with a temporal convolution}
%%%-----------------------------------------------------------------
Let's first define a layer of hetero-synaptic spiking neurons $\postsynaddr \in \postsynaddrspace$ by first describing how each neuron connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses, and each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. For each neuron $\postsynaddr$, we thus define a set of $\Nsyn^\postsynaddr$ synapses, as  $\synapse^\postsynaddr = \{(\presynaddr^\postsynaddr_\ranksyn, \synapticweight^\postsynaddr_\ranksyn, \synapticdelay^\postsynaddr_\ranksyn)\}_{\ranksyn \in [1,\Nsyn^\postsynaddr]}$, where each synapse $\synapse^\postsynaddr_\ranksyn$ is associated to a weight $\synapticweight^\postsynaddr_\ranksyn$, a delay $\synapticdelay^\postsynaddr_\ranksyn$ and a presynaptic address $\presynaddr^\postsynaddr_\ranksyn$. This defines the full connectivity of the hetero-synaptic model.

@@

The activation function of our spiking neuron is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
It transforms this list of weights into a probability with the following formula:
$
Pr(k=\postsynaddr \; \vert \; \timev) =
\frac 1 Z
{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
$ 
where $\mathcal{C}^\postsynaddr(\timev) = \sum
\activeweights^\postsynaddr(t)
$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%
In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the hetero-synaptic delays model. 
%
\subsection{Task definition: fast motion detection}
%%%-----------------------------------------------------------------
Let us now define a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the field of vision. In animals with a fovea, this is particularly useful as it allows to move the area with the highest density of photoreceptors in the environment, for example at a point of interest. For example, saccades are rapid movements of the eyeball that reposition the center of vision. In humans, these are very frequent (on average one every 3 per second~\citep{dandekar_neural_2012}). They are produced very rapidly (about $80~\ms$) and display a whole range of speeds. At a more microscopic scale, the human gaze moves with an incessant drift similar to a Brownian-like trajectory~\citep{poletti_head-eye_2015}. To maintain the full generality of the task, we will define eye movements using a form of random walk~\citep{engbert_integrated_2011}. This approach first defines a finite set of possible 2D motions in polar coordinates. Based on the distribution of biological motions, we simplified it by selecting a set of eye movements as the Cartesian product of N-V-phi=8 linearly spaced motion directions and N-Y-velocity=7 different velocities (see inset of \seeFig{motion_task}-(a)). Note that the velocities are sampled on a geometric scale between $1/4$ and $v-speed_base = 4$. Next, we define one trajectory of gaze as segments whose duration is drawn from a Poisson distribution with an average block length = $20~\ms$ - similar to a Lévy flight~\citep{}. Finally, the trajectory is integrated by assuming 1/ that velocities are uniformly and independently sampled from the set of different motion-sets and 2/ that motion is uniform during a time segment. The resulting instances yield trajectories qualitatively similar to those observed for human eye movements (\seeFig{motion_task}-(a)).

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \note{TODO: redo figure according to caption}
    \caption{
    {\bf Motion detection task.} {\bf (a)} To mimic the effect of a saccadic-like movement, the axis of view moves following a step-like random walk, and we show here on example path of the trajectory. {\bf (b)} We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view (here the cross at the center of the image) such as to reproduce the effect of displacing the eye. {\bf (c)} Recording the dynamics of the sub-image as a function of time, it generates a naturalistic movie which may be transformed to an event-based representation. Mimicking the retina, this representation codes for proportional increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. This will constitute the input to the spiking neural network.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (see \seeFig{motion_task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration N-T = $250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a movement of one pixel per frame (i.e., per time bin).

@@

To transform each movie into events, we compute a gradient image (initialized at zero) by adding the gradient of the pixels' intensity over two successive frames. If, on a specific pixel at that specific timestamp, the absolute value of this gradient exceeds a threshold, an event is generated. The event has either an OFF or ON polarity, respectively whether the gradient is negative or positive. This signed threshold value is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is as a consequence similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, that is, a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF) and time $\timev_\arank$  (\seeFig{motion_task}-(c)). The goal here is to infer the correct motion solely by observing these events. 

@@

%: fig:kernels

@@

    %\vspace{-3cm}

@@

%
%%%-----------------------------------------------------------------
%: fig:accuracy
%%%-----------------------------------------------------------------
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/accuracy.pdf}
    \caption{Accuracy as a function of the number of computation load for the hetero-synaptic delays model (in blue) and for a method using 2D time surfaces (in red). The relative computational load (on a log axis) is controlled by changing the percentage of active weights relative to the dense convolution kernel. We observe a similar accuracy than HOTS, yet that our model could achieve a similar accuracy with significantly fewer coefficients.}
    \label{fig:accuracy}
\end{figure}
%%%-----------------------------------------------------------------
%
\subsection{Testing with natural-like textures}
%%%-----------------------------------------------------------------
To test our model, we will quantify its ability to categorize different motions. In that order, we use a set of synthetic visual stimuli, \textit{Motion Clouds}~\citep{leon_motion_2012} which are natural-like random textures for which we can control for velocity, among other parameters (\seeFig{motionclouds}). In particular, we will set the spatial size to $(\Nx, \Ny)$ and consider a discretization of time with a time step of $1$ such that $\timev \in \mathbb{N}$. Movies' duration are here set to $\Ntime=400$.
This procedure defines a set of textures with different spatial properties and different motions $\va{v}_k$ with  $1 \le k \le \Nclass$ and $\Nclass=8$ defined by a constant speed and linearly spaced directions:
% of a motion cloud is determined by a constant $k_v$: with a constant speed $\speed$ (see \fig{motionclouds}-(a)): 
%\begin{equation*}
%%\begin{gathered}
%\va{v}_k = %\binom{\cos(\frac{2\pi\cdot k_v}{N_v})}{\sin(\frac{2\pi\cdot k_v}{N_v})}
%  \begin{pmatrix}
%    \cos(2\pi\cdot \frac{k}{\Nclass})\\
%    \sin(2\pi\cdot \frac{k}{\Nclass})
%  \end{pmatrix}
%%\end{gathered}
%\end{equation*}
$
\speed_\kernelind = 
  ( 
    \speed \cdot \cos(2\pi\cdot \frac{\kernelind}{\Nspeed}),
    \speed \cdot \sin(2\pi\cdot \frac{\kernelind}{\Nspeed})
  )
$ (\seeFig{motionclouds}-(a) for an illustration).
For any given velocity, we also varied the parameters of the textures, such as the mean and variance of the orientation or spatial frequency content. This method provides a rich dataset of textured movies for which we know the ground truth for motion.
%
%%%-----------------------------------------------------------------
%: fig:motion_clouds
%%%-----------------------------------------------------------------
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_clouds.pdf}
    \caption{{\bf Motion detection task.} {\bf (a)} The motion direction represented as the plain red vector, other possible motion directions are represented in light red. {\bf (b)} A screenshot of one generated naturalistic textured stimulus at a specific time. % $\timev_\arank$. 
    {\bf (c)} The corresponding ON (in red) and OFF (in blue) event stream generated from the stimuli on (b) and constituting the input to the spiking neural network.}
    \label{fig:motion_clouds}
\end{figure}
%%%-----------------------------------------------------------------
